---
title: "Running PFDA on FAUST outputs"
author: 
- Evan Greene
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FAUST PFDA}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  dpi = 300
)
```

# Introduction

This vignette provides information about how to apply PFDA to the annotated cout matrix produced by FAUST. 

```{r libraries_and_sim, message = FALSE, echo=FALSE}
# load libraries and code to simulate data.
suppressPackageStartupMessages({
  library(mvtnorm)
  library(MASS)
  library(flowCore)
  library(flowWorkspace)
  library(ggplot2)
  library(cowplot)
  library(knitr)
  library(dplyr)
  library(tidyr)
  library(faust)
  library(ggridges)
})

Posdef <- function (n, ev = runif(n, 1, 2))
{
    #function written by Ravi Varadhan, from r-help mailing list
    #Thu Feb 7, 20:02:30 CET 2008
    #Generates a positive definine matrix of dimension n
    #ev bounds the variance by 2
    Z <- matrix(ncol=n, rnorm(n^2))
    decomp <- qr(Z)
    Q <- qr.Q(decomp)
    R <- qr.R(decomp)
    d <- diag(R)
    ph <- d / abs(d)
    O <- Q %*% diag(ph)
    Z <- t(O) %*% diag(ev) %*% O
    return(Z)
}

labelMeanVector <- function(meanVector) {
    baseStr <- paste0("V",seq(length(meanVector)))
    tokenStr <- rep("+",length(meanVector))
    tokenStr[which(meanVector==0)] <- "-"
    return(paste0(paste0(baseStr,tokenStr),collapse=""))
}

genClusterCentersWithLabels <- function(possibleCenterMat,
                                        nPop=5,
                                        seedVal=0)
{
    pcMat <- t(possibleCenterMat)
    allPops <- expand.grid(split(pcMat,rep(seq(nrow(pcMat)),ncol(pcMat))))
    if (nPop > nrow(allPops)) {
        print("Too many clusters relative to specMat/possibleCenterMat.")
        stop("Reduce number of clusters or increase possible mean vectors.")
    }
    clusterIndex <- sample(seq(nrow(allPops)),nPop)
    outMat <- allPops[clusterIndex,,drop=FALSE]
    outLab <- as.character(apply(outMat,1,labelMeanVector))
    outList <- list(fixedMeanMatrix=outMat,fixedLabelVector=outLab)
    return(outList)
}

getSimTrueCountMatrix <- function(truthList)
{
    uniqueNames <- c()
    for (i in seq(length(truthList))) {
        uniqueNames <- append(uniqueNames,names(truthList[[i]]))
    }
    uniqueNames <- sort(unique(uniqueNames))
    truthMat <- matrix(0,nrow=length(truthList),ncol=length(uniqueNames))
    colnames(truthMat) <- uniqueNames
    rownames(truthMat) <- names(truthList)
    for (i in seq(length(truthList))) {
        cv <- truthList[[i]]
        for (cName in colnames(truthMat)) {
            lookup <- which(names(cv) == cName)
            if (length(lookup)) {
                truthMat[i,cName] <- cv[lookup]
            }
        }
    }
    return(truthMat)
}

simSample <- function(sampleDim,
                      sampleSize,
                      transformationType,
                      mixtureType="gaussianOnly",
                      fixedMeanMatrix=NA,
                      fixedLabelVector=NA,
                      noiseDim=0,
                      probVecSample,
                      isKnockout=FALSE,
                      isSpikeIn=FALSE,
                      sRegime=0,
                      targetRanks=c(length(probVecSample)-1,length(probVecSample)),
                      hasBatchEffect=FALSE,
                      batchEffect=0
                      )
{
    numClusters <- nrow(fixedMeanMatrix)
    probVec <- probVecSample
    knockoutStatus <- "No knockout"
    if (isKnockout) {
        targetProbs <- sort(probVec)
        targetProb <- targetProbs[ceiling(length(targetProbs)/2)]
        targetLookup <- which(probVec == targetProb)
        modProbLookups <- setdiff(seq(length(probVec)),targetLookup)
        knockoutStatus <- fixedLabelVector[targetLookup]
        probVec[modProbLookups] <- (probVec[modProbLookups] + (targetProb/length(modProbLookups)))
        probVec[targetLookup] <- 0
    }
    if ((sRegime) && (isSpikeIn)) {
        currentSpikeMass <- probVec[targetRanks[2]]
        targetSpikeMass <- probVec[targetRanks[1]]
        massIncrement <- currentSpikeMass/(length(probVec)-1)
        probVec <- probVec + massIncrement
        probVec[targetRanks[2]] <- 0
        probVec <- probVec - (targetSpikeMass * probVec)
        probVec[targetRanks[2]] <- targetSpikeMass
        if (abs(sum(probVec) - 1) > 1e-9) {
            print(probVec)
            print(sum(probVec))
            print(abs(sum(probVec) - 1))
            stop("Error in probability reapportionment")
        }
        if (min(probVec) == probVec[targetRanks[2]]) {
            stop("Error in spiking in population")
        }
    }
    sampleSizeVec <- as.vector(t(rmultinom(1,sampleSize,probVec)))
    outData <- matrix(nrow=0,ncol=(sampleDim+noiseDim))
    outLabel <- c()
    for (clusterNumber in seq(numClusters)) {
        if (sampleSizeVec[clusterNumber] == 0) {
            next
        }
        currentMu <- as.numeric(fixedMeanMatrix[clusterNumber,,drop=TRUE])
        if (hasBatchEffect) {
            currentMu <- currentMu + batchEffect 
        }
        subjectShift <- round(rnorm(length(currentMu),mean=0,sd=(1/sqrt(2))))
        currentMu <- currentMu + subjectShift 
        currentLabel <- fixedLabelVector[clusterNumber]
        outLabel <- append(outLabel,rep(currentLabel,sampleSizeVec[clusterNumber]))
        currentSigma<- Posdef(sampleDim)
        if (mixtureType == "tPlusGauss") {
            if ((clusterNumber %% 2) == 0) {
                currentSample <- mvrnorm(sampleSizeVec[clusterNumber],mu=currentMu,Sigma=currentSigma)
            }
            else {
                currentSample <- rmvt(sampleSizeVec[clusterNumber],delta=currentMu,sigma=currentSigma,df=2)
            }
        }
        else if (mixtureType == "tOnly") {
            currentSample <- rmvt(sampleSizeVec[clusterNumber],delta=currentMu,sigma=currentSigma,df=2)
        }
        else  {
            currentSample <- mvrnorm(sampleSizeVec[clusterNumber],mu=currentMu,Sigma=currentSigma)
        }
        if (is.vector(currentSample)) {
            currentSample <- t(as.matrix(currentSample))
        }
        if (noiseDim > 0) {
            currentSigma<- Posdef(noiseDim)
            noiseSample <- mvrnorm(nrow(currentSample),mu=rep(5,noiseDim),Sigma=currentSigma)
            if (is.vector(noiseSample)) {
                noiseSample <- t(as.matrix(noiseSample))
            }
            currentSample <- cbind(currentSample,noiseSample)
        }
        outData <- rbind(outData,t(apply(currentSample,1,transformationType)))
    }
    colnames(outData) <- paste0("V",seq(ncol(outData)))
    outList <- list(sampleMatrix=outData,sampleLabels=outLabel,knockoutInfo=knockoutStatus)
    return(outList)
}

simulateExperiment <- function(
                               meanVectorBoundsMatrix, 
                               numSamples=100, 
                               randomSeed=0,
                               transformationList=list(function(x){return(x)},
                                                       function(x){return(x)},
                                                       function(x){return(x)}), 
                               noiseDimension=5,
                               probVecIn, 
                               minSampleSize=5000,
                               maxSampleSize=40000,
                               tncp=10000,
                               knockoutNumber=0, 
                               spikeInFlag=FALSE, 
                               targetRanks=c(length(probVecIn)-1,length(probVecIn)),
                               useBatchFlag=FALSE, 
                               batchEffectShift=0.75,
                               fixedResFlag=FALSE, 
                               responderStatusVec=c(NA)
                               )
{
    probVecForSim <- sort(probVecIn,decreasing=TRUE)
    sampleSpecs <- genClusterCentersWithLabels(
        possibleCenterMat=meanVectorBoundsMatrix,
        nPop=length(probVecForSim),
        seedVal=randomSeed
    )
    currentTransformation <- transformationList[[1]]
    currentSample <- 1
    startKnockoutNum <- numSamples - knockoutNumber + 1
    isKnockoutSample <- FALSE
    regime <- rep(0,numSamples)
    regime[sample(seq(numSamples),(numSamples/2))] <- 1
    if (fixedResFlag) {
        regime <- rep(0,numSamples)
        regime[which(responderStatusVec==1)] <- 1
    }
    nextBatchEffect <- rep((-1*batchEffectShift),ncol(sampleSpecs$fixedMeanMatrix))
    regimeList <- knockoutList <- labelsList <- flowList <- truthList <- list()
    while (currentSample <= numSamples) {
        nextSampleSize <- min(max(minSampleSize,round(rt(1,df=3,ncp=tncp))),maxSampleSize)
        nextRegime <- regime[currentSample]
        if (currentSample >= startKnockoutNum) {
            isKnockoutSample <- TRUE
        }
        else {
            isKnockoutSample <- FALSE
        }
        if ((useBatchFlag) && (((currentSample - 1) %% 10) == 0)) {
            nextBatchEffect <- nextBatchEffect + batchEffectShift
            if ((currentSample - 1) > floor(numSamples/3)) {
                currentTransformation <- transformationList[[2]]
            }
            if ((currentSample - 1) > floor((2*(numSamples/3)))) {
                currentTransformation <- transformationList[[3]]
            }
        }

        sampleData <- simSample(
            sampleDim=ncol(sampleSpecs$fixedMeanMatrix),
            sampleSize=nextSampleSize,
            transformationType=currentTransformation,
            fixedMeanMatrix=sampleSpecs$fixedMeanMatrix,
            fixedLabelVector=sampleSpecs$fixedLabelVector,
            noiseDim=noiseDimension,
            probVecSample=probVecForSim,
            isKnockout=isKnockoutSample,
            isSpikeIn=spikeInFlag,
            sRegime=nextRegime,
            targetRanks=targetRanks,
            hasBatchEffect=useBatchFlag,
            batchEffect=nextBatchEffect
        )
        if (currentSample < 10) {
            outName <- paste0("sample00",currentSample)
        }
        else if (currentSample < 100) {
            outName <- paste0("sample0",currentSample)
        }
        else {
            outName <- paste0("sample",currentSample)
        }
        ff <- flowCore::flowFrame(sampleData$sampleMatrix)
        flowList <- append(flowList,ff)
        names(flowList)[length(flowList)] <- outName
        labelsList <- append(labelsList,list(sampleData$sampleLabels))
        names(labelsList)[length(labelsList)] <- outName
        truthList <- append(truthList,list(table(sampleData$sampleLabels)))
        names(truthList)[length(truthList)] <- outName
        knockoutList <- append(knockoutList,list(table(sampleData$knockoutInfo)))
        names(knockoutList)[length(knockoutList)] <- outName
        regimeList <- append(regimeList,list(nextRegime))
        names(regimeList)[length(regimeList)] <- outName
        currentSample <- currentSample + 1
    }
    truthMat <- getSimTrueCountMatrix(truthList)
    
    outputList <- list(
        "flowFrameList"=flowList,
        "truthValueList"=truthList,
        "truthMat"=truthMat,
        "labelsList"=labelsList,
        "knockoutList"=knockoutList,
        "spikedPop"=sampleSpecs$fixedLabelVector[targetRanks[2]],
        "spikedInPopList"=regimeList
    )
    return(outputList)
}
sampleSize <- 2500
ssLB <- 2500
ssUB <- 2500
dataTransformations <- list(function(x){return((x))},
                                function(x){return((x))},
                                function(x){return((x))})
startProb <- 0.1425
refClusterProbVec <- c(
    startProb,rep(startProb/2,2),rep(startProb/4,4),rep(startProb/8,8),rep(startProb/16,16),rep(startProb/32,32),rep(startProb/64,64)
)
numberOfClusters <- 25
clusterProbVec <- sort(refClusterProbVec[seq(numberOfClusters)],decreasing=TRUE)
if (sum(clusterProbVec) < 1) {
    residualPV <- 1-sum(clusterProbVec)
    clusterProbVec <- (clusterProbVec + (residualPV/length(clusterProbVec)))
}
#
#dataset 1 -- make a gating set to illustrate depth-score tuning
#
dimension <- 
specMat <- matrix(0,nrow=2,ncol=dimension)
specMat[2,] <- c(10,10,10,10,10,10,10,10)
simmedExp1 <- simulateExperiment(
    meanVectorBoundsMatrix=specMat,
    numSamples=10,
    transformationList=dataTransformations,
    noiseDimension=0,
    probVecIn=clusterProbVec,
    minSampleSize=ssLB,
    maxSampleSize=ssUB,
    randomSeed=12345,
    tncp=sampleSize
)

specMat[2,] <- c(3.5,3.5,3.5,3.5,3.5,3.5,3.5,3.5)
simmedExp2 <- simulateExperiment(
    meanVectorBoundsMatrix=specMat,
    numSamples=10,
    transformationList=dataTransformations,
    noiseDimension=0,
    probVecIn=clusterProbVec,
    minSampleSize=ssLB,
    maxSampleSize=ssUB,
    randomSeed=12345,
    tncp=sampleSize
)
ffList <- append(simmedExp1[["flowFrameList"]],
                 simmedExp2[["flowFrameList"]])
names(ffList) <- paste0("Sample_",sprintf('%0.2d', 1:length(ffList)))
fs <- as(ffList, "flowSet")
gs <- flowWorkspace::GatingSet(fs)
```

# Annotation thresholds - depth score tuning

We beging by simulating a dataset.

```{r at01, message = FALSE, echo=TRUE, eval=FALSE}
#
#generate the annotation thresholds 
#
set.seed(123)
projPath <- file.path(tempdir(),"FAUST")
if (dir.exists(projPath)) {
   unlink(projPath, recursive=TRUE)
}
dir.create(projPath, recursive = TRUE)
generateAnnotationThresholds(
    gatingSet           = gs,
    startingCellPop     = "root",
    projectPath         = projPath,
    plottingDevice      = "png"
)
```
```{r at01a, message = FALSE, echo=FALSE, eval = TRUE, results='hide'}
set.seed(123)
projPath <- file.path(tempdir(),"FAUST")
if (dir.exists(projPath)) {
   unlink(projPath, recursive=TRUE)
}
dir.create(projPath, recursive = TRUE)
invisible(generateAnnotationThresholds(
    gatingSet           = gs,
    startingCellPop     = "root",
    projectPath         = projPath,
    plottingDevice      = "png"
))
```

We next load the counts into the analysis and derive a parent count. To fit a PFDA model, we will assign a reponder status
to each sample against which we can model.

```{r prepare_counts, eval=FALSE, echo=TRUE}
count_df <- as.data.frame(readRDS(file.path(projPath,"faustData","faustCountMatrix.rds")))
cellPops <- setdiff(colnames(count_df),c(0_0_0_0_0))
count_df$parentCount <- apply(count_df,1,sum)
count_df$subjectID <- paste0("Sample_",sprintf('%0.2d', 1:20))
count_df$modelRT <- c(rep(1,10),rep(0,10))
```

To apply PFDA, we must have a biological compartment in mind whose sub-phenotypes we wish to investigate.
Here are simulating generic data, and so will assume phenotypes that are `V1+`, `V4-`, and `V5+` represent
a biological compartment. T

```{r multivariate model, eval=FALSE, echo=TRUE}
pfdaExampleCellPops <- cellPops[grepl("V1\\+",cellPops)]
pfdaExampleCellPops <- pfdaExampleCellPops[grepl("V4-",pfdaExampleCellPops)]
pfdaExampleCellPops <- pfdaExampleCellPops[grepl("V5\\+",pfdaExampleCellPops)]
pfdaExampleGlmerDFPrep <- count_df[,c(pfdaExampleCellPops,"parentCount","modelRT","subjectID")]
pfdaExampleGlmerDF <- gather(pfdaExampleGlmerDFPrep,key=cellPop,value=count,-c("parentCount","modelRT","subjectID"))
pfdaExampleGlmerDF$popName <- as.factor(paste0("Pop_",as.numeric(as.factor(pfdaExampleGlmerDF$cellPop))))
pfdaExampleGlmerDF$obsFactor <- as.factor(paste0("Obs_",seq(nrow(pfdaExampleGlmerDF))))
```

We now fit the multivariate PFDA model using a binomial GLMM.

```{r pfda_model, eval=FALSE, echo=TRUE}
pfdaExampleGlmerFit <- glmer(
    formula=cbind(count,(parentCount-count))  ~ popName*modelRT + (1|obsFactor),
    data=pfdaExampleGlmerDF,
    family="binomial",
    nAGQ=0,
    control=glmerControl(
        optimizer="bobyqa",
        optCtrl = list(maxfun = 1e9),
        boundary.tol = 1e-9,
        tolPwrss=1e-9,
        check.conv.singular = .makeCC(action = "warning",  tol = 1e-12)
    )
)
```

Having fit the model, we conduct a test of increased abundance using `glht`. 

```{r linear_combo, echo=TRUE, eval=FALSE}
pfdaExampleCoefNames = names(fixef(pfdaExampleGlmerFit))
pfdaExampleGLHTMat <- matrix(0, nrow = 1, ncol = length(pfdaExampleCoefNames))
pfdaExampleGLHTMat[,which(pfdaExampleCoefNames == "modelRT1")] <- 1
pfdaExampleGLHTMat[1,which(pfdaExampleCoefNames %like% ":modelRT1")] <- 1/(length(which(pfdaExampleCoefNames%like%":modelRT1"))+1)
rownames(pfdaExampleGLHTMat) <- c("Phenotype_Compartment")
pfdaExampleMulti <- glht(pfdaExampleGlmerFit,pfdaExampleGLHTMat, alternative = "greater")
pfdaExampleObsPval <- round(summary(pfdaExampleMulti)$test$pvalues[1],9)
pfdaExampleCT95 <- confint(pfdaExampleMulti,level=0.95)
pfdaExampleMultiResults <- as.data.frame(pfdaExampleCT95$confint)
pfdaExampleMultiResults$upr <- 3
pfdaExampleMultiResults$Compartment <- rownames(pfdaExampleMultiResults)
pfdaExampleMultiResults$ModelType <- "PFDA"
pfdaExampleMultiResults$Component <- "PFDA"
```