---
title: "Introduction to FAUST R package"
author: 
- Evan Greene
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{FAUST Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  fig.width = 6,
  fig.height = 12,
  dpi = 100,
  fig.align = "center"
)
```

# FAUST Overview

Full Annotation Using Shape-constrained Trees (FAUST) is a method to perform discovery and annotation of phenotypes in single-cell data from flow and mass cytometry experiments.

This implementation targets the gating set data structure in the [`flowWorkspace`](https://bioconductor.org/packages/release/bioc/html/flowWorkspace.html) Bioconductor package.

When applied to a cytometry study stored in a gating set, the `faust` R function creates a directory called "faustData" located at the setting of a parameter called "projectPath".

The principal output is an **annotated count matrix** `faustCountMatrix.rds` which is written to the "faustData" directory.

- **Rows** in the count matrix correspond to **samples** collected in the single cell experiment.
- The **columns** in the count matrix are **cell populations** discovered by the pipeline. 
- The **columns** are *annotated* by a selected subset of *markers* used in conducting the experiment.
- The **column annotations** define, in terms of these *markers*, the **phenotypes** of all **cell populations** discovered by the pipeline.
- **Entries** in the count matrix correspond the **number of cells** in a sample that belong to a discovered cell population. 

This count matrix can be loaded into R using the `readRDS` function.

This vignette provides a brief overview of how to use the implementation provided in the `faust` package.

# An example workflow with simulated data.

```{r libraries_and_sim, message = FALSE, echo=FALSE}
# load libraries and code to simluate data.
suppressPackageStartupMessages({
  library(mvtnorm)
  library(MASS)
  library(flowCore)
  library(flowWorkspace)
  library(ggplot2)
  library(cowplot)
  library(knitr)
  library(dplyr)
  library(tidyr)
  library(faust)
})

Posdef <- function (n, ev = runif(n, 1, 2))
{
    #function written by Ravi Varadhan, from r-help mailing list
    #Thu Feb 7, 20:02:30 CET 2008
    #Generates a positive definine matrix of dimension n
    #ev bounds the variance by 2
    Z <- matrix(ncol=n, rnorm(n^2))
    decomp <- qr(Z)
    Q <- qr.Q(decomp)
    R <- qr.R(decomp)
    d <- diag(R)
    ph <- d / abs(d)
    O <- Q %*% diag(ph)
    Z <- t(O) %*% diag(ev) %*% O
    return(Z)
}

labelMeanVector <- function(meanVector) {
    baseStr <- paste0("V",seq(length(meanVector)))
    tokenStr <- rep("+",length(meanVector))
    tokenStr[which(meanVector==0)] <- "-"
    return(paste0(paste0(baseStr,tokenStr),collapse=""))
}

genClusterCentersWithLabels <- function(possibleCenterMat,
                                        nPop=5,
                                        seedVal=0)
{
    pcMat <- t(possibleCenterMat)
    allPops <- expand.grid(split(pcMat,rep(seq(nrow(pcMat)),ncol(pcMat))))
    if (nPop > nrow(allPops)) {
        print("Too many clusters relative to specMat/possibleCenterMat.")
        stop("Reduce number of clusters or increase possible mean vectors.")
    }
    clusterIndex <- sample(seq(nrow(allPops)),nPop)
    outMat <- allPops[clusterIndex,,drop=FALSE]
    outLab <- as.character(apply(outMat,1,labelMeanVector))
    outList <- list(fixedMeanMatrix=outMat,fixedLabelVector=outLab)
    return(outList)
}

getSimTrueCountMatrix <- function(truthList)
{
    uniqueNames <- c()
    for (i in seq(length(truthList))) {
        uniqueNames <- append(uniqueNames,names(truthList[[i]]))
    }
    uniqueNames <- sort(unique(uniqueNames))
    truthMat <- matrix(0,nrow=length(truthList),ncol=length(uniqueNames))
    colnames(truthMat) <- uniqueNames
    rownames(truthMat) <- names(truthList)
    for (i in seq(length(truthList))) {
        cv <- truthList[[i]]
        for (cName in colnames(truthMat)) {
            lookup <- which(names(cv) == cName)
            if (length(lookup)) {
                truthMat[i,cName] <- cv[lookup]
            }
        }
    }
    return(truthMat)
}

simSample <- function(sampleDim,
                      sampleSize,
                      transformationType,
                      mixtureType="gaussianOnly",
                      fixedMeanMatrix=NA,
                      fixedLabelVector=NA,
                      noiseDim=0,
                      probVecSample,
                      isKnockout=FALSE,
                      isSpikeIn=FALSE,
                      sRegime=0,
                      targetRanks=c(length(probVecSample)-1,length(probVecSample)),
                      hasBatchEffect=FALSE,
                      batchEffect=0
                      )
{
    numClusters <- nrow(fixedMeanMatrix)
    probVec <- probVecSample
    knockoutStatus <- "No knockout"
    if (isKnockout) {
        targetProbs <- sort(probVec)
        targetProb <- targetProbs[ceiling(length(targetProbs)/2)]
        targetLookup <- which(probVec == targetProb)
        modProbLookups <- setdiff(seq(length(probVec)),targetLookup)
        knockoutStatus <- fixedLabelVector[targetLookup]
        probVec[modProbLookups] <- (probVec[modProbLookups] + (targetProb/length(modProbLookups)))
        probVec[targetLookup] <- 0
    }
    if ((sRegime) && (isSpikeIn)) {
        currentSpikeMass <- probVec[targetRanks[2]]
        targetSpikeMass <- probVec[targetRanks[1]]
        massIncrement <- currentSpikeMass/(length(probVec)-1)
        probVec <- probVec + massIncrement
        probVec[targetRanks[2]] <- 0
        probVec <- probVec - (targetSpikeMass * probVec)
        probVec[targetRanks[2]] <- targetSpikeMass
        if (abs(sum(probVec) - 1) > 1e-9) {
            print(probVec)
            print(sum(probVec))
            print(abs(sum(probVec) - 1))
            stop("Error in probability reapportionment")
        }
        if (min(probVec) == probVec[targetRanks[2]]) {
            stop("Error in spiking in population")
        }
    }
    sampleSizeVec <- as.vector(t(rmultinom(1,sampleSize,probVec)))
    outData <- matrix(nrow=0,ncol=(sampleDim+noiseDim))
    outLabel <- c()
    for (clusterNumber in seq(numClusters)) {
        if (sampleSizeVec[clusterNumber] == 0) {
            next
        }
        currentMu <- as.numeric(fixedMeanMatrix[clusterNumber,,drop=TRUE])
        if (hasBatchEffect) {
            currentMu <- currentMu + batchEffect 
        }
        subjectShift <- round(rnorm(length(currentMu),mean=0,sd=(1/sqrt(2))))
        currentMu <- currentMu + subjectShift 
        currentLabel <- fixedLabelVector[clusterNumber]
        outLabel <- append(outLabel,rep(currentLabel,sampleSizeVec[clusterNumber]))
        currentSigma<- Posdef(sampleDim)
        if (mixtureType == "tPlusGauss") {
            if ((clusterNumber %% 2) == 0) {
                currentSample <- mvrnorm(sampleSizeVec[clusterNumber],mu=currentMu,Sigma=currentSigma)
            }
            else {
                currentSample <- rmvt(sampleSizeVec[clusterNumber],delta=currentMu,sigma=currentSigma,df=2)
            }
        }
        else if (mixtureType == "tOnly") {
            currentSample <- rmvt(sampleSizeVec[clusterNumber],delta=currentMu,sigma=currentSigma,df=2)
        }
        else  {
            currentSample <- mvrnorm(sampleSizeVec[clusterNumber],mu=currentMu,Sigma=currentSigma)
        }
        if (is.vector(currentSample)) {
            currentSample <- t(as.matrix(currentSample))
        }
        if (noiseDim > 0) {
            currentSigma<- Posdef(noiseDim)
            noiseSample <- mvrnorm(nrow(currentSample),mu=rep(5,noiseDim),Sigma=currentSigma)
            if (is.vector(noiseSample)) {
                noiseSample <- t(as.matrix(noiseSample))
            }
            currentSample <- cbind(currentSample,noiseSample)
        }
        outData <- rbind(outData,t(apply(currentSample,1,transformationType)))
    }
    colnames(outData) <- paste0("V",seq(ncol(outData)))
    outList <- list(sampleMatrix=outData,sampleLabels=outLabel,knockoutInfo=knockoutStatus)
    return(outList)
}

simulateExperiment <- function(
                               meanVectorBoundsMatrix, 
                               numSamples=100, 
                               randomSeed=0,
                               transformationList=list(function(x){return(x)},
                                                       function(x){return(x)},
                                                       function(x){return(x)}), 
                               noiseDimension=5,
                               probVecIn, 
                               minSampleSize=5000,
                               maxSampleSize=40000,
                               tncp=10000,
                               knockoutNumber=0, 
                               spikeInFlag=FALSE, 
                               targetRanks=c(length(probVecIn)-1,length(probVecIn)),
                               useBatchFlag=FALSE, 
                               batchEffectShift=0.75,
                               fixedResFlag=FALSE, 
                               responderStatusVec=c(NA)
                               )
{
    probVecForSim <- sort(probVecIn,decreasing=TRUE)
    sampleSpecs <- genClusterCentersWithLabels(
        possibleCenterMat=meanVectorBoundsMatrix,
        nPop=length(probVecForSim),
        seedVal=randomSeed
    )
    currentTransformation <- transformationList[[1]]
    currentSample <- 1
    startKnockoutNum <- numSamples - knockoutNumber + 1
    isKnockoutSample <- FALSE
    regime <- rep(0,numSamples)
    regime[sample(seq(numSamples),(numSamples/2))] <- 1
    if (fixedResFlag) {
        regime <- rep(0,numSamples)
        regime[which(responderStatusVec==1)] <- 1
    }
    nextBatchEffect <- rep((-1*batchEffectShift),ncol(sampleSpecs$fixedMeanMatrix))
    regimeList <- knockoutList <- labelsList <- flowList <- truthList <- list()
    while (currentSample <= numSamples) {
        nextSampleSize <- min(max(minSampleSize,round(rt(1,df=3,ncp=tncp))),maxSampleSize)
        nextRegime <- regime[currentSample]
        if (currentSample >= startKnockoutNum) {
            isKnockoutSample <- TRUE
        }
        else {
            isKnockoutSample <- FALSE
        }
        if ((useBatchFlag) && (((currentSample - 1) %% 10) == 0)) {
            nextBatchEffect <- nextBatchEffect + batchEffectShift
            if ((currentSample - 1) > floor(numSamples/3)) {
                currentTransformation <- transformationList[[2]]
            }
            if ((currentSample - 1) > floor((2*(numSamples/3)))) {
                currentTransformation <- transformationList[[3]]
            }
        }

        sampleData <- simSample(
            sampleDim=ncol(sampleSpecs$fixedMeanMatrix),
            sampleSize=nextSampleSize,
            transformationType=currentTransformation,
            fixedMeanMatrix=sampleSpecs$fixedMeanMatrix,
            fixedLabelVector=sampleSpecs$fixedLabelVector,
            noiseDim=noiseDimension,
            probVecSample=probVecForSim,
            isKnockout=isKnockoutSample,
            isSpikeIn=spikeInFlag,
            sRegime=nextRegime,
            targetRanks=targetRanks,
            hasBatchEffect=useBatchFlag,
            batchEffect=nextBatchEffect
        )
        if (currentSample < 10) {
            outName <- paste0("sample00",currentSample)
        }
        else if (currentSample < 100) {
            outName <- paste0("sample0",currentSample)
        }
        else {
            outName <- paste0("sample",currentSample)
        }
        ff <- flowCore::flowFrame(sampleData$sampleMatrix)
        flowList <- append(flowList,ff)
        names(flowList)[length(flowList)] <- outName
        labelsList <- append(labelsList,list(sampleData$sampleLabels))
        names(labelsList)[length(labelsList)] <- outName
        truthList <- append(truthList,list(table(sampleData$sampleLabels)))
        names(truthList)[length(truthList)] <- outName
        knockoutList <- append(knockoutList,list(table(sampleData$knockoutInfo)))
        names(knockoutList)[length(knockoutList)] <- outName
        regimeList <- append(regimeList,list(nextRegime))
        names(regimeList)[length(regimeList)] <- outName
        currentSample <- currentSample + 1
    }
    truthMat <- getSimTrueCountMatrix(truthList)
    
    outputList <- list(
        "flowFrameList"=flowList,
        "truthValueList"=truthList,
        "truthMat"=truthMat,
        "labelsList"=labelsList,
        "knockoutList"=knockoutList,
        "spikedPop"=sampleSpecs$fixedLabelVector[targetRanks[2]],
        "spikedInPopList"=regimeList
    )
    return(outputList)
}
dimension <- 5 
sampleSize <- 500 
ssLB <- 500 
ssUB <- 500
specMat <- matrix(0,nrow=2,ncol=dimension)
specMat[2,] <- c(8,8,8,8,8)
dataTransformations <- list(function(x){return((x))},
                                function(x){return((x))},
                                function(x){return((x))})
startProb <- 0.1425
refClusterProbVec <- c(
    startProb,rep(startProb/2,2),rep(startProb/4,4),rep(startProb/8,8),rep(startProb/16,16),rep(startProb/32,32),rep(startProb/64,64)
)
numberOfClusters <- 5
clusterProbVec <- sort(refClusterProbVec[seq(numberOfClusters)],decreasing=TRUE)
if (sum(clusterProbVec) < 1) {
    residualPV <- 1-sum(clusterProbVec)
    clusterProbVec <- (clusterProbVec + (residualPV/length(clusterProbVec)))
}
```

After installing and loading the `faust` library, the `R` function 
<code style="color:skyblue">faust</code> is available to analyze cytometry data.

This function takes as input a `GatingSet` data structure.
This data structure can be constructed using the
[`flowWorkspace`](https://bioconductor.org/packages/release/bioc/html/flowWorkspace.html) Bioconductor package.
The gating set should contain all the experimental samples you wish to analyze using <code style="color:skyblue">faust</code>.

Here, we construct a gating set using simulated data to demonstrate how to use the <code style="color:skyblue">faust</code> function. 
The gating set contains 10 simulated sample, each of 1000 observations.
The samples are simulated from a multivariate Gaussian mixture with 5 components.
By design, this means there are 5 distinct clusters or "phenotypes" in the gating set.

```{r simulate_gatingset, warning=FALSE, eval=TRUE, echo=TRUE}
simmedExp <- simulateExperiment(
    meanVectorBoundsMatrix=specMat,
    numSamples=5,
    transformationList=dataTransformations,
    noiseDimension=0,
    probVecIn=clusterProbVec,
    minSampleSize=ssLB,
    maxSampleSize=ssUB,
    randomSeed=12345,
    tncp=sampleSize
)
fs <- as(simmedExp[["flowFrameList"]], "flowSet")
gs <- flowWorkspace::GatingSet(fs)
```

## Using **<span style="color:skyblue">faust</span>** function in **R**

Once cytometry data is stored in a gating set, you can apply the **<span style="color:skyblue">faust</span>** function to the gating set.
You will need to identify the cell population in the gating set to analyze, and a directory on your computer that will store the results.

Since we simulated the data set, we will apply the **<span style="color:skyblue">faust</span>** function to the "root" population,
which comprises the all observations in the data set.
In real cytometry data, the **<span style="color:skyblue">faust</span>** function is usually applied to a pre-gated population such as lymphocytes.

We will store the results in a temporary directory created at the `projPath`.
By setting the `annotationsApproved` parameter to `TRUE`, we run the method from end-to-end.
Additional **<span style="color:skyblue">faust</span>** parameters can be read about by typing `?faust` at the R prompt.

```{r run_faust_on_simulated_data, warning=FALSE, eval=TRUE, echo=TRUE}
projPath <- file.path(tempdir(),"FAUST")
dir.create(projPath, recursive = TRUE)

faust(
    gatingSet           = gs,
    startingCellPop     = "root",
    projectPath         = projPath,
    annotationsApproved = TRUE
)
```

## Examining output

Once the **<span style="color:skyblue">faust</span>** function completes running, the primary output is the annotated count matrix.

```{r show_count_matrix, eval=TRUE, echo=TRUE}
count_df <- as.data.frame(readRDS(file.path(projPath,"faustData","faustCountMatrix.rds")))
count_df
```
We see that there are six columns in the annotated count matrix.
Five are phenotypes discovered by `faust` and annotated in terms of the simulated markers.
The sixth column is annotated "0_0_0_0_0" and corresponds to all observations that were assigned a phenotype which
was not observed frequently enough to survive down-selection.

Single cell annotations and visualizations can be constructed using the function `makeAnnotationEmbedding`.
Here we show the annotation embedding across the experiment colored by the `faustLabels`.

```{r generate_annotation_embeddings, eval=TRUE, echo=TRUE, results='asis'}
snVec <- list.files(file.path(projPath,"faustData","sampleData"))
annoEmbed <- faust::makeAnnotationEmbedding(	      
    projectPath=projPath,
    sampleNameVec=snVec
)

p <- ggplot(annoEmbed,aes(x=umapX,y=umapY,color=faustLabels))+
  geom_point()+
  theme_classic()+
  xlab("Annotation Embedding X")+
  ylab("Annotation Embedding Y")+
  facet_grid(vars(sampleOfOrigin))

p
```
We can compare the single cell annotations to the simulated truth.

```{r cross_tab, eval=TRUE, echo=TRUE}
trueLabels <- Reduce(append,simmedExp[["labelsList"]])
faustLabels <- annoEmbed$faustLabels
table(trueLabels,faustLabels)
```

The simulated true labels are along the rows, and the columns are the faust annotations.
We see that the faust labels along the column re-order the markers in terms of the depth score.
